Denotational Semantics of Gradual Typing using Synthetic Guarded Domain Theory

### 2024-07-12 paper-new/concrete-relational-model.tex
tweak adequacy prose to flow better

\eric{Check this}

In this section, we prove the adequacy of the interpretation of term precision.

\begin{comment}
First we establish some notation. Fix a morphism $f : 1 \to \li \mathbb{N} \cong
\li \mathbb{N}$. We write $f \da n$ to mean that there exists $m$ such that $f =
\delta^m(\eta n)$ and $f \da \mho$ to mean that there exists $m$ such that $f =
\delta^m(\mho)$.

Recall that $\ltbisim$ denotes the relation on value morphisms defined as the
closure of the intensional error-ordering on morphisms under weak bisimilarity
on bboth sides. Specialized to the setting of morphisms from $1 \to U \li
\mathbb{N}$, we have $f \ltbisim g$ iff there exist $f'$ and $g'$ with
%
\[ f \bisim_{\li \mathbb{N}} f' \le_{\li \mathbb{N}} g' \bisim_{\li \mathbb{N}} g. \]\max{$\le$? I thought it was $\ltdyn$}
%
We first note that in this ordering, the semantics of error is not equivalent to
the semantics of the diverging term. The main result we would like to show is as
follows: If $f \ltbisim g : \li \mathbb{N}$, then:
\begin{itemize}
  \item If $f \da n$ then $g \da n$.
  \item If $g \da \mho$ then $f \da \mho$.
  \item If $g \da n$ then $f \da n$ or $f \da \mho$.
\end{itemize}
%
However, this is actually not provable! Intuitively, the issue is
that to prove this result we need to the above result is a ``global''\max{what does ``global'' mean?} result, and it is not possible to prove such results
in the guarded setting\max{isn't everything we do in a ``guarded setting''?}. In particular, if we tried to prove the above result in
the guarded setting we would run into a problem where we would have a natural
number ``stuck'' under a $\later$, with no way to get out the underlying number.
%
Thus, to prove our adequacy result, we need to leave the guarded
setting\max{``pass to leave the guarded setting''?}. We have seen in Section \ref{sec:big-step-term-semantics} that clock
quantification makes this possible.
\end{comment}

Recall the big-step term semantics $- \Downarrow -$ defined in Section
\ref{sec:big-step-term-semantics}. The result we seek to show is the following:
% If $f \bisim^{gl} f' \ltls^{gl} g' \bisim^{gl} g$, then

\begin{theorem}[adequacy]
  If $\cdot \vdash M \ltdyn N : \nat$, then:
  \begin{itemize}
    \item If $M \Downarrow n$ then $N \Downarrow n$.
    \item If $N \Downarrow \mho$ then $M \Downarrow \mho$.
    \item If $N \Downarrow n$ then $M \Downarrow n$ or $M \Downarrow \mho$.
  \end{itemize}
\end{theorem}

We have already established graduality above, which states that syntactic term
precision implies that the denoted terms are related in the semantic
interpretation of term precision.
%
% Recall as discussed in that section that the definitions of our model are all
% parameterized by a clock $k$. 
%
It therefore remains to show that the denotation of term precision in our model
$f \ltsqbisim{}{} g$, where $f, g : \li \mathbb{N}$, implies a relationship
between the partial elements $\mathbb{N} + {\mho}$ corresponding to $f$ and $g$. Recall
that to define the big-step semantics we used the equivalence of the
globalization of the free error domain and Capretta's coinductive delay monad:
$\li^{gl} \mathbb{N} \cong \delay(\mathbb{N} + {\mho})$. We now extend this
correspondence to a relationship between a global version of our denotation of
term precision and the corresponding relation on the delay monad.

>>>>

Now that we have established our denotational model, we can, as in
Section~\ref{sec:big-step-term-semantics}, extract a big step semantics. Our final
goal then is to prove that the extensional squares in our model imply
the graduality property for this final big step semantics.
\begin{comment}
First we establish some notation. Fix a morphism $f : 1 \to \li \mathbb{N} \cong
\li \mathbb{N}$. We write $f \da n$ to mean that there exists $m$ such that $f =
\delta^m(\eta n)$ and $f \da \mho$ to mean that there exists $m$ such that $f =
\delta^m(\mho)$.

Recall that $\ltbisim$ denotes the relation on value morphisms defined as the
closure of the intensional error-ordering on morphisms under weak bisimilarity
on bboth sides. Specialized to the setting of morphisms from $1 \to U \li
\mathbb{N}$, we have $f \ltbisim g$ iff there exist $f'$ and $g'$ with
%
\[ f \bisim_{\li \mathbb{N}} f' \le_{\li \mathbb{N}} g' \bisim_{\li \mathbb{N}} g. \]\max{$\le$? I thought it was $\ltdyn$}
%
We first note that in this ordering, the semantics of error is not equivalent to
the semantics of the diverging term. The main result we would like to show is as
follows: If $f \ltbisim g : \li \mathbb{N}$, then:
\begin{itemize}
  \item If $f \da n$ then $g \da n$.
  \item If $g \da \mho$ then $f \da \mho$.
  \item If $g \da n$ then $f \da n$ or $f \da \mho$.
\end{itemize}
%
However, this is actually not provable! Intuitively, the issue is
that to prove this result we need to the above result is a ``global''\max{what does ``global'' mean?} result, and it is not possible to prove such results
in the guarded setting\max{isn't everything we do in a ``guarded setting''?}. In particular, if we tried to prove the above result in
the guarded setting we would run into a problem where we would have a natural
number ``stuck'' under a $\later$, with no way to get out the underlying number.
%
Thus, to prove our adequacy result, we need to leave the guarded
setting\max{``pass to leave the guarded setting''?}. We have seen in Section \ref{sec:big-step-term-semantics} that clock
quantification makes this possible.
\end{comment}
The result we seek to show is the following:
% If $f \bisim^{gl} f' \ltls^{gl} g' \bisim^{gl} g$, then
\begin{theorem}[adequacy]
  If $\cdot \vdash M \ltdyn N : \nat$, then:
  \begin{itemize}
    \item If $M \Downarrow n$ then $N \Downarrow n$.
    \item If $N \Downarrow \mho$ then $M \Downarrow \mho$.
    \item If $N \Downarrow n$ then $M \Downarrow n$ or $M \Downarrow \mho$.
  \end{itemize}
\end{theorem}
%
% Recall as discussed in that section that the definitions of our model are all
% parameterized by a clock $k$. 
%
It therefore remains to show that the denotation of term precision in our model
$f \ltsqbisim{}{} g$, where $f, g : \li \mathbb{N}$, implies a relationship
between the partial elements $\mathbb{N} + {\mho}$ corresponding to $f$ and $g$. Recall
that to define the big-step semantics we used the equivalence of the
globalization of the free error domain and Capretta's coinductive delay monad:
$\li^{gl} \mathbb{N} \cong \delay(\mathbb{N} + {\mho})$. We now extend this
correspondence to a relationship between a global version of our denotation of
term precision and the corresponding relation on the delay monad.
